Learning Features and Abstract Actions for Computing Generalized Plans

### 2018-11-13 doc/camera-ready/paper.tex
add/remove some commas

Generalized planning studies the computation of plans  that   solve   multiple  instances
%\cite{srivastava08learning,bonet09automatic,srivastava:generalized,hu:generalized,BelleL16,anders:generalized}.
\cite{srivastava08learning,bonet:icaps2009,hu:generalized,BelleL16,anders:generalized}.
For example, the plan  that iteratively  picks  a  clear block above $x$ 
and places it  on the table, achieves the goal $clear(x)$
in \emph{any} instance of the Blocksworld where the gripper is initially empty.
Once this  general plan or  policy is derived it can be  applied to solve an infinite collection
of instances that involve different initial states, different objects, and  different (ground)  actions.

In the basic formulation due  to \citeay{hu:generalized}, a generalized plan
is a mapping of observations into actions that are assumed to be common
among  all the instances. More recently, this formulation has been extended by
\citeay{bonet:ijcai2018} to account for relational domains like Blocksworld  where the sets of
objects and  actions change from instance to instance.
In the new formulation, the observations  are replaced by  a set of boolean
and numerical features $F$ and a set of \emph{abstract actions} $A_F$.
These abstract actions are  \emph{sound} and \emph{complete} if they track the
effects  of the actions on  the features in a suitable way.
The resulting generalized plans  map feature values into abstract actions, and soundness
ensures that the application of an abstract action can be mapped back into the application of
a concrete action with the same effect  over the features. Moreover, the form of the abstract actions
ensures that generalized plans can be computed using fully observable non-deterministic (FOND) planners
once the generalized planning problem is transformed into a FOND problem.
%after some transformations.

@@

Bonet and Geffner's generalized planning formulation assumes that the
features and abstract actions are given. In this work we address this limitation
by  \emph{showing  how the features and abstract actions  can be learned from
the primitive predicates used to define the instances and sampled state transitions.}
For example,  the general policy for achieving $clear(x)$ % of \citeay{bonet:ijcai2018}
is obtained using a FOND planner on an abstraction that consists of  a  boolean
feature $H$  that tracks whether the gripper holds a block,  a numerical feature
$n(x)$ that counts the number of blocks above $x$, and two abstract actions:
one with preconditions $\neg H$ and $n(x) > 0$, and effects $H$ and $n(x)\mminus$
(decrement $n(x)$), and the other with precondition $H$ and effect $\neg H$.
In this work, we show how to obtain such policies from STRIPS instances alone,
without having to provide the features and the abstract actions by hand.

The work has relation to a number of research threads in planning, knowledge
representation, and machine learning. We make use of SAT solvers and description
logics for learning features and abstract actions. %\cite{sat-handbook,dl-handbook}.
The abstract actions provide a model from which the plans are obtained via
transformations and FOND planners \cite{geffner:book,ghallab:book}.
The model is not the model of an instance but a generalized model for obtaining
plans that work for multiple instances. In this sense, the work is different than
action model learning \cite{yang:action-learning} and model-based reinforcement
learning and closer in aims to work on learning general policies from examples or
experience \cite{martin-geffner:generalized,fern:bias,mazebase,general-drl}.
Generalized planning has also been formulated as a problem in first-order logic
\cite{srivastava:generalized}, and general plans over finite horizons have been
derived using first-order regression \cite{boutilier2001symbolic,wang2008first,van2012solving,sanner:practicalMDPs}.
Our approach differs from first-order approaches in the use of propositional
planners, and from purely learning approaches, in the formal guarantees that
are characteristic of planning: if the learned abstract actions are  sound,
the resulting general plans must be correct.

@@

A \emph{generalized planning problem} $\Q$ is a collection of planning instances $P$.
An instance $P$ is a  classical planning problem expressed in some compact language
as a tuple $P=\tup{V,I,G,A}$ where $V$ is a set of state variables that can take a
finite set of values (boolean or not), $I$ is a set of atoms over $V$ defining an
initial state $s_0$, $G$ is a set of literals over $V$ describing the goal
states, and $A$ is a set of actions $a$ with their preconditions and effects that
define the set $A(s)$ of actions applicable in any state $s$, and the successor
state function $f(a,s)$, for any state $s$ and $a \in A(s)$.
A state is a valuation over $V$, and a solution to $P$ is an applicable action
sequence $\pi=a_0,\ldots,a_n$ that generates  a state sequence $s_0,s_1,\ldots,s_{n}$
where $s_n$ is a goal state (makes $G$ true). In this sequence, $a_i \in A(s_i)$
and $s_{i+1}=f(a_i,s_i)$ for $i=0, \ldots, n-1$. A state $s$ is reachable in $P$
if $s=s_n$ for one such sequence. A solution to the \emph{generalized problem}
$\Q$ is a solution to all instances $P \in Q$. The form of such solutions is
described below.

@@

\medskip
\begin{example}
  The two abstract actions  $\bar a = \langle \neg H,n(x)>0; H,$ $n(x)\mminus\rangle$ %$\bar{a}=\abst{\neg H,n(x)>0}{H,n(x)\mminus}$
  and $\bar{a}'=\abst{H}{\neg H}$ over the features  $F=\{H,n(x)\}$
  is sound but not complete relative to $\Q_{clear}$. 
  They capture the concrete actions that pick blocks from above $x$ and put them away,
  but do not capture other actions like stacking a block above $x$ or picking up a block that is not above $x$.
\end{example}

@@

While the initial and goal expressions $I_F$ and $G_F$ of the abstract problem can be learned 
from the satisfying assignment $\sigma$ as well, for simplicity, in the examples below, they are
provided by hand.\footnote{$G_F = G_{\sigma}$ may be defined as the DNF formula whose terms 
  correspond to the abstract states over $F_{\sigma}$ that correspond to goal states in
  the sample. For each such term, a new abstract action to reach a new goal literal
  from such a term may be required.
  }
% \footnote{$G_F = G_{\sigma}$ can be defined as the DNF formula
% whose terms  correspond to the abstract states over the selected features in $F_{\sigma}$ 
% that are true in some  goal state of the sample  and false in all non-goal states.}

>>>>

Generalized planning studies the computation of plans  that   solve   multiple  instances
%\cite{srivastava08learning,bonet09automatic,srivastava:generalized,hu:generalized,BelleL16,anders:generalized}.
\cite{srivastava08learning,bonet:icaps2009,hu:generalized,BelleL16,anders:generalized}.
For example, the plan  that iteratively  picks  a  clear block above $x$ 
and places it  on the table, achieves the goal $clear(x)$
in \emph{any} instance of the Blocksworld where the gripper is initially empty.
Once this  general plan or  policy is derived, it can be  applied to solve an infinite collection
of instances that involve different initial states, different objects, and  different (ground)  actions.

In the basic formulation due  to \citeay{hu:generalized}, a generalized plan
is a mapping of observations into actions that are assumed to be common
among  all the instances. More recently, this formulation has been extended by
\citeay{bonet:ijcai2018} to account for relational domains like Blocksworld  where the sets of
objects and  actions change from instance to instance.
In the new formulation, the observations  are replaced by  a set of boolean
and numerical features $F$ and a set of \emph{abstract actions} $A_F$.
These abstract actions are  \emph{sound} and \emph{complete} if they track the
effects  of the actions on  the features in a suitable way.
The resulting generalized plans  map feature values into abstract actions, and soundness
ensures that the application of an abstract action can be mapped back into the application of
a concrete action with the same effect  over the features. Moreover, the form of the abstract actions
ensures that generalized plans can be computed using fully observable non-deterministic (FOND) planners,
once the generalized planning problem is transformed into a FOND problem.
%after some transformations.

@@

Bonet and Geffner's generalized planning formulation assumes that the
features and abstract actions are given. In this work, we address this limitation
by  \emph{showing  how the features and abstract actions  can be learned from
the primitive predicates used to define the instances and from sampled state transitions.}
For example,  the general policy for achieving $clear(x)$ % of \citeay{bonet:ijcai2018}
is obtained using a FOND planner on an abstraction that consists of  a  boolean
feature $H$,  that tracks whether the gripper holds a block,  a numerical feature
$n(x)$ that counts the number of blocks above $x$, and two abstract actions:
one with preconditions $\neg H$ and $n(x) > 0$, and effects $H$ and $n(x)\mminus$
(decrement $n(x)$), and the other with precondition $H$ and effect $\neg H$.
We here show how to obtain such policies from STRIPS instances alone,
without having to provide the features and the abstract actions by hand.

This work relates to a number of research threads in planning, knowledge
representation, and machine learning. We make use of SAT solvers and description
logics for learning features and abstract actions. %\cite{sat-handbook,dl-handbook}.
The abstract actions provide a model from which the plans are obtained via
transformations and FOND planners \cite{geffner:book,ghallab:book}.
The model is not the model of an instance but a generalized model for obtaining
plans that work for multiple instances. In this sense, the work is different than
action model learning \cite{yang:action-learning} and model-based reinforcement
learning and closer in aims to work on learning general policies from examples or
experience \cite{martin-geffner:generalized,fern:bias,mazebase,general-drl}.
Generalized planning has also been formulated as a problem in first-order logic
\cite{srivastava:generalized}, and general plans over finite horizons have been
derived using first-order regression \cite{boutilier2001symbolic,wang2008first,van2012solving,sanner:practicalMDPs}.
Our approach differs from first-order approaches in the use of propositional
planners, and from purely learning approaches in the formal guarantees that
are characteristic of planning: if the learned abstract actions are  sound,
the resulting general plans must be correct.

@@

A \emph{generalized planning problem} $\Q$ is a collection of planning instances $P$.
An instance $P$ is a  classical planning problem expressed in some compact language
as a tuple $P=\tup{V,I,G,A}$ where $V$ is a set of state variables that can take a
finite set of values (boolean or not), $I$ is a set of atoms over $V$ defining an
initial state $s_0$, $G$ is a set of literals over $V$ describing the goal
states, and $A$ is a set of actions $a$ with their preconditions and effects, which
define the set $A(s)$ of actions applicable in any state $s$, and the successor
state function $f(a,s)$, for any state $s$ and $a \in A(s)$.
A state is a valuation over $V$, and a solution to $P$ is an applicable action
sequence $\pi=a_0,\ldots,a_n$ that generates  a state sequence $s_0,s_1,\ldots,s_{n}$
where $s_n$ is a goal state (makes $G$ true). In this sequence, $a_i \in A(s_i)$
and $s_{i+1}=f(a_i,s_i)$ for $i=0, \ldots, n-1$. A state $s$ is reachable in $P$
if $s=s_n$ for one such sequence. A solution to the \emph{generalized problem}
$\Q$ is a solution to all instances $P \in Q$. The form of such solutions is
described below.

@@

\medskip
\begin{example}
  The abstract action set $A_F = \{\bar{a}, \bar{a}'\}$, where $\bar{a} = \langle \neg H,n(x)>0; H,$ $n(x)\mminus\rangle$ %$\bar{a}=\abst{\neg H,n(x)>0}{H,n(x)\mminus}$
  and $\bar{a}'=\abst{H}{\neg H}$ , defined over the above feature set $F=\{H,n(x)\}$,
  is sound but not complete relative to $\Q_{clear}$. 
  $A_F$ captures the concrete actions that pick blocks from above $x$ and put them away,
  but not other actions like stacking a block above $x$ or picking up a block that is not above $x$.
\end{example}

@@

While the initial and goal expressions $I_F$ and $G_F$ of the abstract problem can be learned 
from the satisfying assignment $\sigma$ as well, for simplicity, in the examples below they are
provided by hand.\footnote{$G_F = G_{\sigma}$ may be defined as the DNF formula whose terms 
  correspond to the abstract states over $F_{\sigma}$ that correspond to goal states in
  the sample. For each such term, a new abstract action to reach a new goal literal
  from such a term may be required.
  }
% \footnote{$G_F = G_{\sigma}$ can be defined as the DNF formula
% whose terms  correspond to the abstract states over the selected features in $F_{\sigma}$ 
% that are true in some  goal state of the sample  and false in all non-goal states.}
